---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bedslider

<!-- badges: start -->
[![R build status](https://github.com/EmilHvitfeldt/bedslider/workflows/R-CMD-check/badge.svg)](https://github.com/EmilHvitfeldt/bedslider/actions)
<!-- badges: end -->

The goal of bedslider is to allow calculating metrics in a sliding fashion over .bed files.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("EmilHvitfeldt/bedslider")
```

## Example

bedslider includes a randomized dataset `bed_sample` we will use in examples.

```{r}
library(bedslider)
bed_sample
```

This dataset has chromosome information, start positions and methylation values for different measures in the beta columns.

The `bed_slide()` Allows you to calculate metrics in a sliding fashion around each data point.
The `vars` argument is used to specify the columns that contain the values. If multiple columns are selected then they are converted to a matrix before calculations begin.
The funs argument is a list of functions that take a single matrix as input and returns a single numeric value.
the `.i` argument specifies the position variables that we are sliding over.
`size` determines the size of the window of observations we are including, centered around the point. A `size = 100` would mean all points between 50 units before the point and 50 units after the point.

```{r}
bed_slide(bed_sample,
          vars = list(beta3),
          funs = list(metric_mhic),
          .i = start, size = 100)
```
The functions return the beginning data.frame with a column added on for each function with the name of the function.

You can pass multiple functions in the `funs` list to have multiple metrics calculated at once.
This is more efficient than calculating them one by one.

```{r}
bed_slide(bed_sample,
          vars = list(beta3),
          funs = list(metric_mhic, metric_pwd, metric_sample_var),
          .i = start, size = 100)
```

Using a named list will transfer those names to the output.

```{r}
bed_slide(bed_sample,
          vars = list(beta3),
          funs = list(mhic = metric_mhic,
                      pwd = metric_pwd, 
                      sample_var = metric_sample_var),
          .i = start, size = 100)
```
If you supply multiple columns to `vars` then they are combined in the calculation.

```{r}
bed_slide(bed_sample,
          vars = list(beta1, beta2, beta3, beta4, beta5),
          funs = list(mhic = metric_mhic,
                      pwd = metric_pwd, 
                      sample_var = metric_sample_var),
          .i = start, size = 100)
```
## Calculate metrics one column at a time

bedslider does not have the functionality to calculate metrics one column at a time for multiple columns in the same function call.
but bedslider functions are easily pipeable so you can do comparative analysis using pipes and named lists in the `funs` argument.

```{r}
library(dplyr)

bed_sample %>% 
  bed_slide(vars = list(beta1),
            funs = list(beta1_mhic = metric_mhic, 
                        beta1_pwd = metric_pwd, 
                        beta1_sample_var = metric_sample_var),
            .i = start, size = 100) %>%
  bed_slide(vars = list(beta2),
            funs = list(beta2_mhic = metric_mhic, 
                        beta2_pwd = metric_pwd, 
                        beta2_sample_var = metric_sample_var),
            .i = start, size = 100)
```


## Metric functions

bedslider comes builtin with a couple of optimized functions for calculating metrics on matrices. These functions are all prefixed with `metric_`.

- `metric_mhic()`

Calculates the proportion of values that falls inside an interval, with the defaults being (0.2, 0.8)

```{r}
mat <- matrix(seq(0.1, 0.9, by = 0.1), ncol = 3)
mat
```

```{r}
mean(mat > 0.2 & mat < 0.8)

metric_mhic(mat)
```

- `metric_pwd()`

Calculates the mean of the [distance matrix](https://en.wikipedia.org/wiki/Distance_matrix) using the Manhattan distance.

If multiple columns are present then the calculates are done column-wise and averaged.

```{r}
mat[, 1, drop = FALSE]

metric_pwd(mat[, 1, drop = FALSE])
(abs(0.1 - 0.2) + abs(0.1 - 0.3) + abs(0.2 - 0.3)) / 3

metric_pwd(mat)
```

- `metric_sample_var()`

Calculates the variances sample variance within each column.

If multiple columns are present then the calculates are done column-wise and averaged.

```{r}
var(mat[, 1])

metric_sample_var(mat[, 1, drop = FALSE])

metric_sample_var(mat)
```

## Specify arguments in metric functions

If you want to specify arguments in the metric functions then you need can create a little wrapper function and pass that one in

```{r}
mhic_01_09 <- function(x) metric_mhic(x, lower = 0.1, upper = 0.9)

bed_sample %>% 
  bed_slide(vars = list(beta1),
            funs = list(beta1_mhic = mhic_01_09),
            .i = start, size = 100)
```

or modify the function directly inside the list. If you do it this way make sure to use a named list.

```{r}
bed_sample %>% 
  bed_slide(vars = list(beta1),
            funs = list(beta1_mhic = function(x) metric_mhic(x, 0.1, 0.9)),
            .i = start, size = 100)
```

